---
title: "Impact of homelessness on surgical outcomes"
output:
  html_document:
    toc: true
---

# Background

The following observational analysis has been designed to determine the impact of homelessness on surgical outcomes. To assess this impact, if any, we fit several linear models containing their housing status and other potentially confounding variables to determine whether there is an impact and quantify it.

The St Vincent's Hospital Sydney Coding Office (SVHSCO) provided the data sets relating to these outcomes.

# Methods

Binary logistic regression and linear logistic regression were used to analyse associations between our complication factors as independent variables to each endpoint. Using the outputs from these models, we determine which variables have a significant effect on each outcome. We then remove the non-significant variables from the models to arrive at final models from which we can determine effect sizes for each complication factor.

## Enpoints

The primary efficacy endpoint is:

* Surgical complication as `complication`: logical variable indicating whether the patient had a complication associated with their operation.

and the secondary endpoints include:

* Patient Mortality as `mortality`: a logical variable indicating whether the patient died during their hospital admission.
* Length of Stay as`los` a measurement of in days of how long each patient stayed in hospital.
* ICU Admission as `icu_admission`: a logical variable indicating whether the patient was admitted to ICU.
* ICU Hours as `icu_hrs`: a measurement in hours of how long the patient spent in ICU.
* Discharge at Own Risk as `daor`: a logical variable indicating whether patients discharged themselves at their own risk.

## Complication factors

The variable of interest was:

* Housing Status as `status`: the patient's housing status undergoing surgery.

Other potentially confounding variables considered as part of the analysis were:

* Patient Age as `age`: a measurement of the patient's age in years at the time of admission.
* Patient Sex as `sex`: sex of the patient.
* Drug and Alcohol Disorder as `da_disorder`: whether the patient had a drug and alcohol disorder associated with their admission.
* Mental Illness as `mental_illness`: whether the patient had a drug and a mental illness associated with their admission.

The datasets were exported by SVHSCO to a set of google sheets which was then ingested by this analysis document.

# Environment set-up

First we set up our environment by loading the following packages.

```{r, message=FALSE, warning=FALSE}
# load libraries
library(googlesheets4)
library(googledrive)
library(tidyverse)
library(reshape2)
library(janitor)
library(skimr)
library(scales)
library(ggplot2)
library(ggpubr)
library(chisq.posthoc.test)
```

# Data load, clean and feature engineering

The data is contained as a google sheet. The following routine fetches, cleans and generates features from these datasets to create an analysis data frame called `homeless_data`.

The cleaning process involves the following processes:

* All non-surgical patient admissions were filtered out of the data set.
* `NA` values for `assoc_diseases` were encoded as `NONE`
* `NA` values for `icu_hrs` were encoded as 0.

The features we generate are:

* `complication`: a logical variable as defined by the inclusion of `assoc_diseases` beginning with `T8`.
* `icu_admission`: logical variable defined by the `icu_hrs` variable greater than 0.
* `da_disorder`: a logical variable defined by the inclusion of `assoc_diseases` beginning with `F1`, except `F17`.
* `mental_illness`: a logical variable defined by the inclusion of `assoc_diseases` beginning with `F2`, `F3` or `F4`.
* `doar`: as logical variable as defined by a `discharge_status` equal to `Discharge at Own Risk.`

```{r, message=FALSE, warning=FALSE}
# load data by looping through the sheets
sheets <- vector(mode = "list", length = 2)

for (s in 1:2){
  
  sheets[[s]] <- drive_get("Homeless Data") %>%
    read_sheet(sheet = s) %>%
    clean_names() %>%
    select(sex, age, assoc_diseases, spec, los, icu_hrs, discharge_status) %>%
    mutate(status = c("domiciled", "homeless")[s])
  
}

# bind the sheets together and clean
homeless_data <- bind_rows(sheets) %>%
  filter(spec %in% c("OT1", "OT2", "CRS", "NS", "HNO", "PLS", "CTS", "VAS", "SOC", "ENT", "URO", "GIT")) %>%
  mutate_at(vars(c("sex", "spec", "discharge_status", "status")), as.factor) %>%
  replace_na(list(assoc_diseases = "NONE",
                  icu_hrs = 0))

# feature engineering
homeless_data <- homeless_data %>%
  mutate(complication = ifelse(str_detect(assoc_diseases, "T8"), T, F),
         mortality = ifelse(str_detect(as.character(discharge_status), "DEATH"), T, F),
         icu_admission = ifelse(icu_hrs > 0, T, F),
         da_disorder = str_detect(assoc_diseases, paste(paste0("F", c(10:16, 18:19)), collapse = "|")),
         mental_illness = case_when(
           str_detect(assoc_diseases, "F2") ~ T,
           str_detect(assoc_diseases, "F3") ~ T,
           str_detect(assoc_diseases, "F4") ~ T,
           TRUE ~ F
           ),
         daor = ifelse(discharge_status == "Discharge at Own Risk", T, F))

# add in descriptions for spec
homeless_data <- homeless_data %>%
  mutate(spec_desc = case_when(
    spec == 'CRS'                           ~ 'Colorectal Surgery',
    spec == 'CTS'                           ~ 'Cardiothoracic Surgery',
    spec == 'ENT'                           ~ 'Ear, Nose and Throat',
    spec %in% c('GIT', 'CRS', 'HNO', 'SOC') ~ 'General Surgery',
    spec == 'NS'                            ~ 'Neurosurgery',
    spec %in% c('OT1', 'OT2')               ~ 'Orthopaedic Surgery',
    spec == 'PLS'                           ~ 'Plastic and Reconstructive Surgery',
    spec == 'URO'                           ~ 'Urology',
    spec == 'VAS'                           ~ 'Vascular Surgery',
    TRUE                                    ~ as.character(NA)
  ) %>% as.factor())

# skim set by status group
homeless_data %>%
  group_by(status) %>%
  skim()
```

# Results

## Primary Efficacy Endpoint - Surgical complication

To determine the effects on complication risk during surgery we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
comp_glm <- glm(complication ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

comp_glm_summary <- comp_glm %>% summary
comp_glm_summary
```

We observe that the variable `status` (P = `r comp_glm_summary$coefficients["statushomeless", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
comp_glm_final <- glm(complication ~ age + sex + da_disorder + mental_illness, 
                      family = binomial(),
                      data = homeless_data)

comp_glm_final_summary <- comp_glm_final %>% summary
comp_glm_final_summary
```

From this final model we can determine that that:

* For each year of age, the risk of surgical complication increases by a factor of `r exp(comp_glm_final$coefficients[["age"]])` (95% CI = `r exp(confint(comp_glm_final)["age",])`, P = `r comp_glm_final_summary$coefficients["age", 4]`).
* Male patients are `r exp(comp_glm_final$coefficients[["sexM"]])` times more likely to experience surgical complication than females (95% CI = `r exp(confint(comp_glm_final)["sexM",])`, P = `r comp_glm_final_summary$coefficients["sexM", 4]`).
* Patients with drug and alcohol disorders are `r exp(comp_glm_final$coefficients[["da_disorderTRUE"]])` times more likely to experience surgical complication than those without (95% CI = `r exp(confint(comp_glm_final)["da_disorderTRUE",])`, P = `r comp_glm_final_summary$coefficients["da_disorderTRUE", 4]`).
* Patients with associated mental illness are `r exp(comp_glm_final$coefficients[["mental_illnessTRUE"]])` times more likely to experience surgical complication than those without (95% CI = `r exp(confint(comp_glm_final)["mental_illnessTRUE",])`, P = `r comp_glm_final_summary$coefficients["mental_illnessTRUE", 4]`).

## Secondary Enpoints

### Mortality

To determine the effects on mortality risk during surgery we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
mort_glm <- glm(mortality ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

mort_glm_summary <- mort_glm %>% summary
mort_glm_summary
```

We observe that the variables `status` (P = `r mort_glm_summary$coefficients["statushomeless", 4]`), `da_disorder` (P = `r mort_glm_summary$coefficients["da_disorderTRUE", 4]`) and `mental_illness` (P = `r mort_glm_summary$coefficients["mental_illnessTRUE", 4]`) don't contribute to the model. As such we eliminate these terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
mort_glm_final <- glm(mortality ~ age + sex, 
                      family = binomial(),
                      data = homeless_data)

mort_glm_final_summary <- mort_glm_final %>% summary
mort_glm_final_summary
```

From this final model we can determine that that:

* For each year of age, the risk of mortality increases by `r exp(mort_glm_final$coefficients[["age"]])` times (95% CI = `r exp(confint(mort_glm_final)["age",])`, P = `r mort_glm_final_summary$coefficients["age", 4]`).
* Male patients are `r exp(mort_glm_final$coefficients[["sexM"]])` times more likely to experience mortality than females (95% CI = `r exp(confint(mort_glm_final)["sexM",])`, P = `r mort_glm_final_summary$coefficients["sexM", 4]`).

### Length of Stay

To determine the effects on a patient's length of stay, we fit linear models to our dataset. Firstly, we need to make a log transform to our outcome variable to fit the model correctly.

```{r, message=FALSE, warning=FALSE}
# new log transformed outcome variable
homeless_data <- homeless_data %>%
  mutate(los_trans = log(los))

los_lm <- lm(los_trans ~ status + age + sex + da_disorder + mental_illness,
             data = homeless_data)

los_lm_summary <- los_lm %>% summary
los_lm_summary
```

We observe that the variable `sex` (P = `r los_lm_summary $coefficients["sexM", 4]`) didn't contribute to the model. As such we eliminate the terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
los_lm_final <- lm(los_trans ~ status + age + da_disorder + mental_illness,
                   data = homeless_data)

los_lm_final_summary <- los_lm_final %>% summary
los_lm_final_summary
```

From this final model we can determine that that:

* Homeless patients are expected to have stay `r exp(los_lm_final$coefficients[["statushomeless"]])` times longer in hospital than domiciled patients (95% CI = `r exp(confint(los_lm_final)["statushomeless",])`, P = `r los_lm_final_summary$coefficients["statushomeless", 4]`).
* For each year increase of patients age, the length of stay is expected to increase by `r exp(los_lm_final$coefficients[["age"]])` times (95% CI = `r exp(confint(los_lm_final)["age",])`, P = `r los_lm_final_summary$coefficients["age", 4]`).
* Patients with drug and alcohol disorders are expected to have stay `r exp(los_lm_final$coefficients[["da_disorderTRUE"]])` times longer in hospital than those without (95% CI = `r exp(confint(los_lm_final)["da_disorderTRUE",])`, P = `r los_lm_final_summary$coefficients["da_disorderTRUE", 4]`).
* Patients with associated mental illness are expected to have stay `r exp(los_lm_final$coefficients[["mental_illnessTRUE"]])` times longer in hospital than those without (95% CI = `r exp(confint(los_lm_final)["mental_illnessTRUE",])`, P = `r los_lm_final_summary$coefficients["mental_illnessTRUE", 4]`).

### ICU Admission

To determine the effects on ICU admission risk during surgery, we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
icu_glm <- glm(icu_admission ~ status + age + sex + da_disorder + mental_illness, 
               family = binomial(),
               data = homeless_data)

icu_glm_summary <- icu_glm %>% summary
icu_glm_summary
```

We observe that the variable `status` (P = `r mort_glm_summary $coefficients["statushomeless", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
icu_glm_final <- glm(icu_admission ~ age + sex + da_disorder + mental_illness,
                     family = binomial(),
                     data = homeless_data)

icu_glm_final_summary <- icu_glm_final %>% summary
icu_glm_final_summary
```

From this final model we can determine that that:

* For each year of age, the risk of ICU admission increases by `r exp(icu_glm_final$coefficients[["age"]])` times (95% CI = `r exp(confint(icu_glm_final)["age",])`, P = `r icu_glm_final_summary$coefficients["age", 4]`).
* Male patients are `r exp(icu_glm_final$coefficients[["sexM"]])` times more likely to experience ICU admission than females (95% CI = `r exp(confint(icu_glm_final)["sexM",])`, P = `r icu_glm_final_summary$coefficients["sexM", 4]`).
* Patients with drug and alcohol disorders are `r exp(icu_glm_final$coefficients[["da_disorderTRUE"]])` times more likely to experience ICU admission than those without (95% CI = `r exp(confint(icu_glm_final)["da_disorderTRUE",])`, P = `r icu_glm_final_summary$coefficients["da_disorderTRUE", 4]`).
* Patients with associated mental illnesses are `r exp(icu_glm_final$coefficients[["mental_illnessTRUE"]])` times more likely to experience ICU admission than those without (95% CI = `r exp(confint(icu_glm_final)["mental_illnessTRUE",])`, P = `r icu_glm_final_summary$coefficients["mental_illnessTRUE", 4]`).

### ICU Hours

To determine the effects on a patient's length of stay in ICU, we fit linear models to our data set. Firstly, we need to make a log transform to our outcome variable to fit the model properly.

```{r, message=FALSE, warning=FALSE}
homeless_data <- homeless_data %>%
  mutate(icu_hrs_trans = ifelse(icu_hrs == 0, NA, log(icu_hrs)))

icu_lm <- lm(icu_hrs_trans ~ status + age + sex + da_disorder + mental_illness, 
             data = homeless_data)

icu_lm_summary <- icu_lm %>% summary
icu_lm_summary
```

We observe that the variables `status` (P = `r icu_lm_summary$coefficients["statushomeless", 4]`)
`age` (P = `r icu_lm_summary$coefficients["age", 4]`) and `sex` (P = `r icu_lm_summary$coefficients["sexM", 4]`) didn't contribute to the model. As such we eliminate the terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
icu_lm_final <- lm(icu_hrs_trans ~ da_disorder + mental_illness, 
                   data = homeless_data)

icu_lm_final_summary <- icu_lm_final %>% summary
icu_lm_final_summary
```

From this final model we can determine that that:

* Patients with drug and alcohol disorders are `r exp(icu_lm_final$coefficients[["da_disorderTRUE"]])` times longer in ICU admission than those without (95% CI = `r exp(confint(icu_lm_final)["da_disorderTRUE",])`, P = `r icu_lm_final_summary$coefficients["da_disorderTRUE", 4]`).
* Patients with associated mental illnesses are `r exp(icu_lm_final$coefficients[["mental_illnessTRUE"]])` times more likely to experience ICU admission than those without (95% CI = `r exp(confint(icu_lm_final)["mental_illnessTRUE",])`, P = `r icu_lm_final_summary$coefficients["mental_illnessTRUE", 4]`).

### Discharge At Own Risk

To determine the effects on patients risk to discharge themselves at their risk post-surgery, we fit a logistic regression model to our data set as so:

```{r, message=FALSE, warning=FALSE}
daor_glm <- glm(daor ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

daor_glm_summary <- daor_glm %>% summary
daor_glm_summary
```

We observe that the variable `mental_illness` (P = `r daor_glm_summary$coefficients["mental_illnessTRUE", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
daor_glm_final <- glm(daor ~ status + age + sex + da_disorder, 
                      family = binomial(),
                      data = homeless_data)

daor_glm_final_summary <- daor_glm_final %>% summary
daor_glm_final_summary
```

From this final model we can determine that that:

* Homeless patients are `r exp(daor_glm$coefficients[["statushomeless"]])` times more likely to discharge themselves at their own risk than domiciled patients (95% CI = `r exp(confint(daor_glm)["statushomeless",])`, P = `r daor_glm_summary$coefficients["statushomeless", 4]`).
* For each year of age, the likelihood of discharge themselves at their own risk decreases by `r exp(daor_glm$coefficients[["age"]])` (95% CI = `r confint(daor_glm)["age",]`, P = `r daor_glm_summary$coefficients["age", 4]`).
* Male patients are `r exp(daor_glm$coefficients[["sexM"]])` more likely to discharge themselves at their own risk than females (95% CI = `r exp(confint(daor_glm)["sexM",])`, P = `r daor_glm_summary$coefficients["sexM", 4]`).
* Patients with drug and alcohol disorders are `r exp(daor_glm$coefficients[["da_disorderTRUE"]])` more likely to discharge themselves at their own risk than those without (95% CI = `r confint(daor_glm)["da_disorderTRUE",]`, P = `r daor_glm_summary$coefficients["da_disorderTRUE", 4]`).

# Prevalence of confounders amongst population groups

## Continuous confounders

```{r, message=FALSE, warning=FALSE}
# display summary statistics of age across both populations
homeless_data %>%
  group_by(status) %>%
  summarise(ave_age = mean(age),
            median_age = median(age),
            sd_age = sd(age),
            q0 = quantile(age, 0),
            q25 = quantile(age, 0.25),
            q75 = quantile(age, 0.75),
            q100 = quantile(age, 1))

# graph the distributions
p1 <- homeless_data %>%
  ggplot(aes(x = status, y = age, fill = status)) + 
    geom_violin(trim = FALSE) +
    geom_boxplot(width = 0.1, fill = "white") +
    labs(title = "", x = "", y = "Age (years)") +
    theme_classic() +
    theme(legend.position="none")

p1

# conduct test to find whether medians are different
kruskal.test(age ~ status, data = homeless_data)
```


## Categorical confounders

```{r, message=FALSE, warning=FALSE}
# find the contingency tables of confounders amongst both populations
con_tab <- homeless_data %>%
  select(status, sex, da_disorder,  mental_illness) %>%
  lapply(., function(x) table(.$status, x)) %>% .[-1]

con_tab

p2 <- homeless_data %>%
  mutate(sex = ifelse(sex == 'M', T, F)) %>%
  group_by(status) %>%
  summarise_at(vars(sex, da_disorder,  mental_illness), function(x) mean(x)) %>%
  melt(variable.name = "confounder", value.name = "proportion") %>%
  ggplot(aes(x = confounder, y = proportion, fill = status)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = '',
         fill = 'Housing status') +
    ylab('Proportion of population') +
    xlab('') +
    scale_x_discrete(labels = c("sex" = "Male", "da_disorder" = "Drug and alcohol disorder", "mental_illness" = "Mental illness")) +
    theme_classic()

p2

# show these contingency tables as prop tables
con_tab %>% lapply(., function(x) prop.table(x, margin = 1)) 

# determine whether the odds of their occurrence is different using fisher test
con_tab %>% lapply(., fisher.test)
```

```{r, message=FALSE, warning=FALSE}
# combine the graphs together
ggarrange(p1, p2,
          widths = c(0.5, 1),
          common.legend = T,
          legend = "bottom")
```



# Types of surgery given to the two populations

```{r, message=FALSE, warning=FALSE}
# generate contingency tables for observed surgeries between groups
surg_tab <- homeless_data %>%
  with(table(status, spec_desc))

surg_tab

# show this as bar chat of proportions
surg_tab %>% prop.table(., margin = 1) %>% 
  as.data.frame() %>%
  ggplot(aes(x = spec_desc, y = Freq, fill = status)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = 'Distribution of surgical types between different housing statuses',
         fill = 'Housing status') +
  
    ylab('Proportion of surgical cases') +
    xlab('Surgery Type') +
    coord_flip() + theme_classic()

# chi square test to see if the distributions are different
surg_tab %>% chisq.test()

# post hoc analysis showing where the differences are
surg_tab %>% chisq.posthoc.test()
```


# Conclusions

By fitting sets of models to our outcome variables, we have determined that homelessness only affects a patient's length of stay and risk of discharge at own risk. All other analysed outcomes are associated with other confounding variables that are often associated with homelessness itself.
