---
title: "Impact of homelessness on surgical outcomes - VAS"
output:
  html_document:
    toc: true
---

# Background

The following observational analysis has been designed to determine the impact of homelessness on surgical outcomes. To assess this impact, if any, we fit several linear models containing their housing status and other potentially confounding variables to determine whether there is an impact and quantify it.

The St Vincent's Hospital Sydney Coding Office (SVHSCO) provided the data sets relating to these outcomes.

# Methods

Binary logistic regression and linear logistic regression were used to analyse associations between our complication factors as independent variables to each endpoint. Using the outputs from these models, we determine which variables have a significant effect on each outcome. We then remove the non-significant variables from the models to arrive at final models from which we can determine effect sizes for each complication factor.

## Enpoints

The primary efficacy endpoint is:

* Surgical complication as `complication`: logical variable indicating whether the patient had a complication associated with their operation.

and the secondary endpoints include:

* Patient Mortality as `mortality`: a logical variable indicating whether the patient died during their hospital admission.
* Length of Stay as`los` a measurement of in days of how long each patient stayed in hospital.
* ICU Admission as `icu_admission`: a logical variable indicating whether the patient was admitted to ICU.
* ICU Hours as `icu_hrs`: a measurement in hours of how long the patient spent in ICU.
* Discharge against medical advice as `daor`: a logical variable indicating whether patients discharged themselves at their own risk.

## Complication factors

The variable of interest was:

* Housing Status as `status`: the patient's housing status undergoing surgery.

Other potentially confounding variables considered as part of the analysis were:

* Patient Age as `age`: a measurement of the patient's age in decades at the time of admission.
* Patient Sex as `sex`: sex of the patient.
* Drug and Alcohol Disorder as `da_disorder`: whether the patient had a drug and alcohol disorder associated with their admission.
* Mental Illness as `mental_illness`: whether the patient had a drug and a mental illness associated with their admission.

The datasets were exported by SVHSCO to a set of google sheets which was then ingested by this analysis document.

# Environment set-up

First we set up our environment by loading the following packages.

```{r, message=FALSE, warning=FALSE}
# load libraries
library(googlesheets4)
library(googledrive)
library(tidyverse)
library(reshape2)
library(janitor)
library(skimr)
library(scales)
library(ggplot2)
library(ggpubr)
library(chisq.posthoc.test)
library(jtools)
```

# Data load, clean and feature engineering

The data is contained as a google sheet. The following routine fetches, cleans and generates features from these datasets to create an analysis data frame called `homeless_data`.

The cleaning process involves the following processes:

* All non-surgical patient admissions were filtered out of the data set.
* `NA` values for `assoc_diseases` were encoded as `NONE`
* `NA` values for `icu_hrs` were encoded as 0.

The features we generate are:

* `complication`: a logical variable as defined by the inclusion of `assoc_diseases` beginning with `T8`.
* `icu_admission`: logical variable defined by the `icu_hrs` variable greater than 0.
* `da_disorder`: a logical variable defined by the inclusion of `assoc_diseases` beginning with `F1`, except `F17`.
* `mental_illness`: a logical variable defined by the inclusion of `assoc_diseases` beginning with `F2`, `F3` or `F4`.
* `daor`: as logical variable as defined by a `discharge_status` equal to `Discharge at Own Risk` meaning that the patient discharged themselves against medical advice.

```{r, message=FALSE, warning=FALSE}
# load data by looping through the sheets
sheets <- vector(mode = "list", length = 2)

for (s in 1:2){
  
  sheets[[s]] <- drive_get("Homeless Data") %>%
    read_sheet(sheet = s) %>%
    clean_names() %>%
    select(sex, age, assoc_diseases, spec, los, icu_hrs, discharge_status) %>%
    mutate(status = c("domiciled", "homeless")[s])
  
}

# bind the sheets together and clean
homeless_data <- bind_rows(sheets) %>%
  filter(spec %in% c("OT1", "OT2", "CRS", "NS", "HNO", "PLS", "CTS", "VAS", "SOC", "ENT", "URO", "GIT")) %>%
  mutate_at(vars(c("sex", "spec", "discharge_status", "status")), as.factor) %>%
  replace_na(list(assoc_diseases = "NONE",
                  icu_hrs = 0))

# feature engineering
homeless_data <- homeless_data %>%
  mutate(complication = ifelse(str_detect(assoc_diseases, "T8"), T, F),
         age = age/10, # want to represent age in decades in analysis
         mortality = ifelse(str_detect(as.character(discharge_status), "DEATH"), T, F),
         icu_admission = ifelse(icu_hrs > 0, T, F),
         da_disorder = str_detect(assoc_diseases, paste(paste0("F", c(10:16, 18:19)), collapse = "|")),
         mental_illness = case_when(
           str_detect(assoc_diseases, "F2") ~ T,
           str_detect(assoc_diseases, "F3") ~ T,
           str_detect(assoc_diseases, "F4") ~ T,
           TRUE ~ F
           ),
         daor = ifelse(discharge_status == "Discharge at Own Risk", T, F))

# add in descriptions for spec
homeless_data <- homeless_data %>%
  mutate(spec_desc = case_when(
    spec %in% c('GIT', 'CRS', 'HNO', 'SOC') ~ 'GS',
    spec %in% c('OT1', 'OT2')               ~ 'ORT',
    TRUE                                    ~ as.character(spec)
  ) %>% as.factor())

# only focus on vascular patients
homeless_data <- homeless_data %>%
  filter(spec == "VAS")

# skim set by status group
homeless_data %>%
  group_by(status) %>%
  skim()
```

# Results

## Initial contingency table analysis

First, we perform a one-sided Fisher's Exact Test to determine whether an association existed between homelessness and greater surgical complication risk.

```{r, message=FALSE, warning=FALSE}
comp_table <- homeless_data %>%
  with(table(complication, status)) %>%
  .[order(-as.logical(row.names(.))), ]

comp_table

comp_table %>%
  fisher.test(., alternative = 'less')
```

## Prevalence of confounders amongst population groups

### Continuous confounders

```{r, message=FALSE, warning=FALSE}
# display summary statistics of age across both populations
homeless_data %>%
  group_by(status) %>%
  summarise(ave_age = mean(age),
            median_age = median(age),
            sd_age = sd(age),
            q0 = quantile(age, 0),
            q25 = quantile(age, 0.25),
            q75 = quantile(age, 0.75),
            q100 = quantile(age, 1))

# graph the distributions
p1 <- homeless_data %>%
  ggplot(aes(x = status, y = age, fill = status)) + 
    geom_violin(trim = FALSE) +
    geom_boxplot(width = 0.1, fill = "white") +
    labs(title = "", x = "", y = "Age (10 years)") +
    theme_classic() +
    theme(legend.position="none")

p1

# conduct test to find whether medians are different
kruskal.test(age ~ status, data = homeless_data)
```

### Categorical confounders

```{r, message=FALSE, warning=FALSE}
# find the contingency tables of confounders amongst both populations
con_tab <- homeless_data %>%
  select(status, sex, da_disorder,  mental_illness) %>%
  lapply(., function(x) table(.$status, x)) %>% .[-1]

con_tab

p2 <- homeless_data %>%
  mutate(sex = ifelse(sex == 'M', T, F)) %>%
  group_by(status) %>%
  summarise_at(vars(sex, da_disorder,  mental_illness), function(x) mean(x)) %>%
  melt(variable.name = "confounder", value.name = "proportion") %>%
  ggplot(aes(x = confounder, y = proportion, fill = status)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = '',
         fill = 'Housing status') +
    ylab('Proportion of population') +
    xlab('') +
    scale_x_discrete(labels = c("sex" = "Male", "da_disorder" = "Drug and alcohol disorder", "mental_illness" = "Mental illness")) +
    theme_classic()

p2

# show these contingency tables as prop tables
con_tab %>% lapply(., function(x) prop.table(x, margin = 1)) 

# determine whether the odds of their occurrence is different using fisher test
con_tab %>% lapply(., fisher.test)
```

```{r, message=FALSE, warning=FALSE}
# combine the graphs together
ggarrange(p1, p2,
          widths = c(0.5, 1),
          common.legend = T,
          legend = "bottom")
```


## Linear Regression analysis

Here we fit sets of linear models to our outcomes to determine how the variables we are investigating relate to our outcomes of interest and use the coefficients of the models to determine the potential risk they carry.

### Surgical complication

To determine the effects on complication risk during surgery we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
comp_glm <- glm(complication ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

comp_glm_summary <- comp_glm %>% summary
comp_glm_summary
```

We observe that the variable `status` (P = `r comp_glm_summary$coefficients["statushomeless", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
comp_glm_final <- glm(complication ~ age + sex + da_disorder + mental_illness, 
                      family = binomial(),
                      data = homeless_data)

# get interesting stuff into a df
comp_glm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
comp_glm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.))))

# transform the coefficients into risk factors
thing <- comp_glm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot1 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('Complication') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot1
```

### Mortality

To determine the effects on mortality risk during surgery we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
mort_glm <- glm(mortality ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

mort_glm_summary <- mort_glm %>% summary
mort_glm_summary
```

We observe that the variables `status` (P = `r mort_glm_summary$coefficients["statushomeless", 4]`), `da_disorder` (P = `r mort_glm_summary$coefficients["da_disorderTRUE", 4]`) and `mental_illness` (P = `r mort_glm_summary$coefficients["mental_illnessTRUE", 4]`) don't contribute to the model. As such we eliminate these terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
mort_glm_final <- glm(mortality ~ age + sex, 
                      family = binomial(),
                      data = homeless_data)

# get interesting stuff into a df
mort_glm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
thing <- mort_glm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot2 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('Mortality') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot2
```

### Length of Stay

To determine the effects on a patient's length of stay, we fit linear models to our dataset. Firstly, we need to make a log transform to our outcome variable to fit the model correctly.

```{r, message=FALSE, warning=FALSE}
# new log transformed outcome variable
homeless_data <- homeless_data %>%
  mutate(los_trans = log(los))

los_lm <- lm(los_trans ~ status + age + sex + da_disorder + mental_illness,
             data = homeless_data)

los_lm_summary <- los_lm %>% summary
los_lm_summary
```

We observe that the variable `sex` (P = `r los_lm_summary $coefficients["sexM", 4]`) didn't contribute to the model. As such we eliminate the terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
los_lm_final <- lm(los_trans ~ status + age + da_disorder + mental_illness,
                   data = homeless_data)

# get interesting stuff into a df
los_lm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
thing <- los_lm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot3 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('Length of stay') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot3
```

### ICU Admission

To determine the effects on ICU admission risk during surgery, we fit a logistic regression model to our dataset as so:

```{r, message=FALSE, warning=FALSE}
icu_glm <- glm(icu_admission ~ status + age + sex + da_disorder + mental_illness, 
               family = binomial(),
               data = homeless_data)

icu_glm_summary <- icu_glm %>% summary
icu_glm_summary
```

We observe that the variable `status` (P = `r mort_glm_summary $coefficients["statushomeless", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
icu_glm_final <- glm(icu_admission ~ age + sex + da_disorder + mental_illness,
                     family = binomial(),
                     data = homeless_data)

# get interesting stuff into a df
icu_glm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
thing <- icu_glm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot4 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('ICU Admission') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot4
```

### ICU Hours

To determine the effects on a patient's length of stay in ICU, we fit linear models to our data set. Firstly, we need to make a log transform to our outcome variable to fit the model properly.

```{r, message=FALSE, warning=FALSE}
homeless_data <- homeless_data %>%
  mutate(icu_hrs_trans = ifelse(icu_hrs == 0, NA, log(icu_hrs)))

icu_lm <- lm(icu_hrs_trans ~ status + age + sex + da_disorder + mental_illness, 
             data = homeless_data)

icu_lm_summary <- icu_lm %>% summary
icu_lm_summary
```

We observe that the variables `status` (P = `r icu_lm_summary$coefficients["statushomeless", 4]`)
`age` (P = `r icu_lm_summary$coefficients["age", 4]`) and `sex` (P = `r icu_lm_summary$coefficients["sexM", 4]`) didn't contribute to the model. As such we eliminate the terms from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
icu_lm_final <- lm(icu_hrs_trans ~ da_disorder + mental_illness, 
                   data = homeless_data)

# get interesting stuff into a df
icu_lm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
icu_lm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.))))

# transform the coefficients into risk factors
thing <- icu_lm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot5 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('ICU Hours') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot5
```

### Discharge against medical advice

To determine the effects on patients risk to discharge themselves at their risk post-surgery, we fit a logistic regression model to our data set as so:

```{r, message=FALSE, warning=FALSE}
daor_glm <- glm(daor ~ status + age + sex + da_disorder + mental_illness, 
                family = binomial(),
                data = homeless_data)

daor_glm_summary <- daor_glm %>% summary
daor_glm_summary
```

We observe that the variable `mental_illness` (P = `r daor_glm_summary$coefficients["mental_illnessTRUE", 4]`) didn't contribute to the model. As such we eliminate the term from the modelling to reach our final model.

```{r, message=FALSE, warning=FALSE}
daor_glm_final <- glm(daor ~ status + age + sex + da_disorder, 
                      family = binomial(),
                      data = homeless_data)

# get interesting stuff into a df
daor_glm_final %>%
  summ(confint = TRUE, digits = 3)

# transform the coefficients into risk factors
thing <- daor_glm_final %>%
  with(cbind(exp(coefficients(.)), exp(confint(.)))) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename("effect_size" = "V1") %>%
  filter(variable != "(Intercept)")

thing

risk_plot6 <- ggplot(thing, aes(x = variable, y = effect_size)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.1) +
  xlab('Discharge against medical advice') +
  ylab('') +
  ylim(0,7) +
  #scale_y_continuous(breaks= 0:7) +
  scale_x_discrete(labels = c("statushomeless" = "Homeless", "age" = "Age (10 years)", "sexM" = "Male", "da_disorderTRUE" = "D&A disorder", "mental_illnessTRUE" = "Mental illness")) +
  coord_flip() +
  theme_minimal()
  
risk_plot6
```

```{r, message=FALSE, warning=FALSE}
# combine the graphs together
risk_plots <- ggarrange(risk_plot1, risk_plot2, risk_plot3, risk_plot4, risk_plot6,
          ncol = 1,
          widths = c(1, 1, 1, 1, 1, 1),
          common.legend = T,
          legend = "bottom")

annotate_figure(risk_plots,
                top = text_grob("Derived risk factors from models"),
                bottom = text_grob("Risk Factor"))
```


## Types of surgery given to the two populations

```{r, message=FALSE, warning=FALSE}
# generate contingency tables for observed surgeries between groups
surg_tab <- homeless_data %>%
  with(table(status, spec_desc))

surg_tab

# show this as bar chat of proportions
surg_tab %>% prop.table(., margin = 1) %>% 
  as.data.frame() %>%
  ggplot(aes(x = spec_desc, y = Freq, fill = status)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = 'Distribution of surgical types between different housing statuses',
         fill = 'Housing status') +
    ylab('Proportion of surgical cases') +
    xlab('Surgery Type') +
    theme_classic()

# chi square test to see if the distributions are different
surg_tab %>% chisq.test()

# post hoc analysis showing where the differences are
surg_tab %>% chisq.posthoc.test()
```

# Conclusions

By fitting sets of models to our outcome variables, we have determined that homelessness only affects a patient's length of stay and risk of Discharge against medical advice. All other analysed outcomes are associated with other confounding variables that are often associated with homelessness itself.
